---
title: "Prediction Assigment over the Weight Lifting Exercises Dataset"
author: "Francisco Alonso"
date: "November 20, 2015"
output: html_document
---

#Summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset) [1].

The 'Random Forest' model was used obtaining high accuracy for prediction over a test set of 20 records. The accuracy obtained is 0.993 under the test sub-set. This was achieved by training the 'Random Forest' model and applying the Cross-Validation method.

#Loading the Data

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

```{r cache=TRUE, eval=F}
  library(ggplot2)
  library(caret)
  library(dplyr)
  library(parallel)
  library(doParallel)
  originalTrainingSet = read.csv("pml-training.csv", header = TRUE)
```

#Data Preprocessing

The variables with missing values were removed along with variables that are not useful for the prediction process, this was done with the objective of improving the prediction outcome and reduce the amount of data present in the sets to decrease the execution time. The training set was divided in a sub training and sub testing set of 60% and 40% of the original training set respectively. The seed was set too for reproduciblity.

```{r cache=TRUE, eval=F}
  set.seed(42)
  
  na_count <-sapply(originalTrainingSet, function(y) sum(length(which(is.na(y) || y == ""))))
  na_count <- data.frame(na_count)

  naCols <-sapply(na_count, function(y) which(y > 0))
  naCols <- data.frame(naCols)

  trainingSet <- select(originalTrainingSet
                        , - naCols[,1]
                        , -X
                        , -user_name
                        , -raw_timestamp_part_1
                        , -raw_timestamp_part_2
                        , -num_window
                        , -new_window
                        , -cvtd_timestamp)
  trainingSet$classe <- as.factor(trainingSet$classe)

  trainIndex <- createDataPartition(trainingSet$classe, p=0.6,list=FALSE);
  training <- trainingSet[trainIndex,]
  testing <- trainingSet[-trainIndex,]
```

#Model Training

The 'Random Forest' ("rf") model was tested and after finding the estimated set of useful variables for the prediction process that allowed an acceptable execution time, the importance of each was calculated to validate the estimate. The "doParallel" package was used to handle the high amount of data and computations performed. To avoid Over-Fitting the Cross Validation method with K=3 is used.

```{r cache=TRUE, eval=F}
  registerDoParallel(makeCluster(2))
  
  model_rf <- train(classe ~ .
                   ,  method="rf"
                   , data = training
                   , allowParallel=T
                   , trControl = trainControl(method = "cv", number = 3)
                   ,importance=TRUE
                   )
```

##Variable Importance

The most important predictor variable from the cross-validated model is the "roll_belt".

```{r cache=TRUE, eval=F}
  varImportance = varImp(model_rf$finalModel)
```

    roll_belt            1242.83123
    pitch_belt            543.99271
    yaw_belt              635.44856
    total_accel_belt       44.74292
    gyros_belt_x           37.61024

##Accuracy of the Model

Making a predction over the testing sub-set the accuracy of the model can be obtained.

```{r eval=F}
  predTestingSubset <- predict(model_rf , testing)
  confusionMatrix(predTestingSubset, testing$classe)
```

    Overall Statistics
                                          
               Accuracy : 0.993           
                 95% CI : (0.9909, 0.9947)
    No Information Rate : 0.2845          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.9911          
    Mcnemar's Test P-Value : NA              
    Statistics by Class:

                     Class: A Class: B Class: C Class: D Class: E
    Sensitivity            0.9991   0.9921   0.9920   0.9844   0.9931
    Specificity            0.9979   0.9986   0.9971   0.9979   0.9998
    Pos Pred Value         0.9946   0.9941   0.9862   0.9891   0.9993
    Neg Pred Value         0.9996   0.9981   0.9983   0.9970   0.9984
    Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
    Detection Rate         0.2842   0.1919   0.1730   0.1614   0.1825
    Detection Prevalence   0.2858   0.1931   0.1754   0.1631   0.1826
    Balanced Accuracy      0.9985   0.9953   0.9945   0.9912   0.9965

The resulting accuracy for this model was 0.993 on the testing sub-set.

#Prediction Test - 20 test cases

Last, the model is used to predict a test set of 20 elements, this file writer function was provided by the course instructors.

```{r eval=F}
  
  pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
      filename = paste0("problem_id_",i,".txt")
      write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
  }
  
  TestSet = read.csv("pml-testing.csv")
  predictionTest <- predict(model_rf, TestSet)
  pml_write_files(as.vector(predictionTest))
```

#References

[1] Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.

Read more: http://groupware.les.inf.puc-rio.br/har#wle_paper_section#ixzz3sCJ45xdG